Lossy compression of DNA sequencing quality data

### 2015-02-02 manuscript/q-compression.tex
Added another review reference, corrected several spelling mistakes in text from original course report.

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little discernable effects on subsequent
analysis results. The four compression strategies in this study were
able to produce similar results to those obtained with uncompressed
quality sequences in terms of quality control, genome assembly, and
alignment of short read to a reference sequence.

@@

Although the quality values are useful for later steps, the trade-off
between storing effectively double the amount of data and the
resulting effects on subsequent analyses has not been
well-documented. Previous literature on sequence data compression has
largely focused on lossless compression of base
calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Giancarlo:2014rw}.
If a reference exists for the biological data, sequences can be
compressed using offsets into the
reference~\cite{Kozanitis:2011kl}. Typical compression methods fail to
utilize the biases inherent in existing sequencing
technologies. Depending on the sequencing technology used, the base
calling is more accurate in the earlier cycles, gradually losing
quality in later cycles~\cite{Kozanitis:2011kl}. To this end, methods
for compressing these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

@@

The last approach we investigated is based on transforming the
original data into a wholly different representation for which
efficient and proven compression algorithms already exist. In this
study, we described a way of reducing the problem of quality
compression to image compression via blockwise quality data
transformation. This approach is different from the others in that
depending on the image compression algorithm used it is not restricted
to lossy techniques.

@@

Nonetheless, fitting a polynomial and reconstructing the quality
values that it would have yielded after decompression still makes some
sense as it alters the properties of the quality data, making it
highly compressible simply due to the smoothing effect it
provides. Therefore the compression ratio and bits per bp values for
polynomial regression throughout this manuscript refer to bzipped
polynomially approximated quality values.

@@

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-order polynomial
regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the
similar contiguity statistics, the different compression methods vary
noticeably in the amount of SNPs. The order of polynomial has an
inverse relationship with the amount of SNPs detected. The good/bad,
profile and PNG methods detected the least amount of SNPs compared to
the reference genome, outperforming the assembly using the original
quality sequences. A more indepth evaulation is needed to determine
whether these compression methods are missing actual SNPs.

It is important to highlight the result that the 4th-order polynomial
regression outperforms the 5th-order in all of the reference-based
metrics, but not the \emph{de novo} LAP metrics. The 5th-order
polynomial regression assembles approximately 2.5 kb more sequence and
uses roughly $1.2\%$ more of the total read set than the 4th-order
(Supplemental Table 1), which may explain the discrepency in LAP
values because the LAP value is influenced by total
coverage~\cite{Ghodsi:2013hb}.

>>>>

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little descernable effects on subsequent
analysis results. The four compression strategies in this study were
able to produce similar results to those obtained with uncompressed
quality sequences in terms of quality control, genome assembly, and
alignment of short read to a reference sequence.

@@

Although the quality values are useful for later steps, the trade-off
between storing effectively double the amount of data and the
resulting effects on subsequent analyses has not been
well-documented. Previous literature on sequence data compression has
largely focused on lossless compression of base
calls~\cite{Giancarlo:2009fk,Nalbantoglu:2010uq,Zhu:2013qr,Deorowicz:2013hq,Giancarlo:2014rw}.
If a reference exists for the biological data, sequences can be
compressed using offsets into the
reference~\cite{Kozanitis:2011kl}. Typical compression methods fail to
utilize the biases inherent in existing sequencing
technologies. Depending on the sequencing technology used, the base
calling is more accurate in the earlier cycles, gradually losing
quality in later cycles~\cite{Kozanitis:2011kl}. To this end, methods
for compressing these quality sequences utilize these general quality
trends. \textsc{SlimGene}~\cite{Kozanitis:2011kl} is one such tool
that builds a fixed-state Markov model on adjacent quality scores and
uses Huffman Coding for compression based on the transition
probabilities.

@@

The last approach we investigated is based on transforming the
original data into a wholly different representation for which
efficient and proven compression algorithms already exist. In this
study, we described a way of reducing the problem of quality
compression to image compression via block-wise quality data
transformation. This approach is different from the others in that
depending on the image compression algorithm used it is not restricted
to lossy techniques.

@@

Nonetheless, fitting a polynomial and reconstructing the quality
values that it would have yielded after decompression still makes some
sense as it alters the properties of the quality data, making it
highly compressible simply due to the smoothing effect it
provides. Therefore the compression ratio and bits per bp values for
polynomial regression throughout this manuscript refer to 
polynomially approximated quality values compressed with zip.

@@

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
sequences. All compression methods other than 0-order polynomial
regression produce an NG50 ranging from 3.18--3.19 Mbps. Despite the
similar contiguity statistics, the different compression methods vary
noticeably in the amount of SNPs. The order of polynomial has an
inverse relationship with the amount of SNPs detected. The good/bad,
profile and PNG methods detected the least amount of SNPs compared to
the reference genome, outperforming the assembly using the original
quality sequences. A more in-depth evaluation is needed to determine
whether these compression methods are missing actual SNPs.

It is important to highlight the result that the 4th-order polynomial
regression outperforms the 5th-order in all of the reference-based
metrics, but not the \emph{de novo} LAP metrics. The 5th-order
polynomial regression assembles approximately 2.5 kb more sequence and
uses roughly $1.2\%$ more of the total read set than the 4th-order
(Supplemental Table 1), which may explain the descrepency in LAP
values because the LAP value is influenced by total
coverage~\cite{Ghodsi:2013hb}.
